Policing for Profit and the Mobilization of Black Voters

### 2021-05-07 markdown/draft_paper.Rmd
FINAL FIRST DRAFT!!

author:
- Jonathan Ben-Menachem^[PhD Student, Columbia University, Department of Sociology (jb4487@columbia.edu)]
- Kevin Morris^[PhD Student, CUNY Graduate Center, Department of Sociology (kmorris@gradcenter.cuny.edu)]
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    keep_tex: true
bibliography: "policing_turnout.bib"
link-citations: yes
fontsize: 12pt
header-includes:
    - \usepackage{rotating}
    - \usepackage{setspace}
abstract: |
  The American criminal legal system is an important site of political socialization: sociologists show how the alienating effect of criminalization pushes people away from public institutions, and political scientists have measured the effects of policing and incarceration on political participation. Despite this burgeoning literature, no research has directly investigated how police ticketing practices affect political participation. We use three approaches to answer this question: national survey data, a city-level analysis using Census of Governments data and the national voter file, and individual-level ticketing data from Hillsborough County, Florida between 2003 and 2019. The survey data show that Black people who report that they were recently stopped by police are more likely to vote compared to those who were not stopped. Nationally, we estimate that each 10 percent increase in city fines and fees collections between 2012 and 2017 is associated with a roughly 0.6 percentage point turnout decrease but was not associated with Black turnout. In Tampa, Florida, we estimate that being ticketed decreases an individual’s likelihood of voting in federal elections by 1.7 percentage points, but depressive effects are smaller for Black residents.

@@

Fines and fees are increasingly recognized as a form of racist revenue extraction connected to marginalized communities’ alienation from government. After Michael Brown was killed by the Ferguson Police Department in 2014, a US Department of Justice investigation into the city’s police and courts demonstrated that the municipality was engaged in a practice that advocates now refer to as “policing for profit.” The city’s reliance on fines and fees to fund government functions grew from 13 to 23 percent of the total budget between fiscal years 2012 and 2015. From 2012 to 2014, the Department of Justice found that 85 percent of vehicle stops, 90 percent of citations, and 93 percent of arrests targeted Black people. By contrast, just two-thirds of Ferguson’s residents are Black [@UnitedStatesDepartmentofJusticeCivilRightsDivision2015].

It wasn’t just a Ferguson problem, or even a Missouri problem. American cities’ reliance on fines and fees revenue increased significantly following the 2008 recession---as local tax revenues dropped and tax increases became less politically viable, jurisdictions increased the amounts of fines and fees and imposed them more frequently in order to fund government services [@Singla2020; @Harris2017].

Work exploring how criminalization directly and indirectly influences political participation has exploded in recent years. Scholars have found that criminal legal contact (i.e. arrest, conviction, incarceration) consistently discourages voting [@Weaver2010; @Burch2011; @White2019a]. Such work has largely focused on the effects of highly disruptive contact with the criminal legal system such as incarceration and felony convictions [@Burch2014; @Lee2014]. This paper is the first to explicitly theorize how local ticketing practices shape the political socialization of ticketed individuals. Additionally, our project represents the first use of individual-level administrative data to identify the causal effect of ticketing on voter behavior. 
The use of administrative data marks an important step forward in our understanding of how low-level contact with the criminal legal system structures political participation. Past work looking at the individual-level effects of low-level contact has relied entirely on survey or qualitative data [e.g. @Walker2014; @Weaver2010]. Existing research allows for the testing of specific psychological mechanisms and personal interpretations of criminal legal contact, but does not allow us to generalize more broadly. As Weaver and Lerman [-@Weaver2010, page 821] note, it may also introduce measurement bias. Our analysis investigates actual voting behavior following actual ticketing events, not reported voting behavior or reported exposure to police ticketing. The administrative data therefore allow us both to sidestep reporting error and to observe the behavior of the 232,568 individuals ticketed over a 6 year period---a far larger pool than even the most robust surveys.

Before turning to the administrative data, we begin with a novel analysis of 2016 and 2020 national survey data regarding voters’ experience with police stops and arrests. We find that Black respondents who reported personal or proximal experience with police stops in the last 12 months are more likely to vote than Black people who did not, while we find no statistically significant relationship between stops and voting among non-Black respondents.

Next, we join fees and fines data from municipalities around the country in 2012 and 2017 with a novel dataset estimating racial turnout for each municipality in the country. While we find that increased fees and fines likely cause lower turnout among non-Black Americans, we find no effect of fees and fines on Black turnout.

@@

The American criminal legal system directly shapes the American electorate and excludes Black voters. Today, more than five million Americans are legally barred from voting due to a felony conviction, and one in every sixteen Black Americans of voting age is disenfranchised [@Uggen2020]. Following the implementation of the Fifteenth Amendment in 1870, over a dozen U.S. states passed laws denying voting rights to people who were convicted of felonies. In their analysis of how racial threat catalyzes felon disenfranchisement, @Behrens2003 found that increases in the share of a state’s prison population that is nonwhite drastically increases the likelihood of legislation permanently disenfranchising people convicted of felonies. As @Shofner1963 notes, felony disenfranchisement has often been tied explicitly to the increased presence of Black Americans in the electorate.

@@

Beyond de jure disenfranchisement, the spatial and racial concentration of criminalization influences affected communities’ perceptions of government. Monica Bell calls this process “legal estrangement,” a conceptual framework meant to capture those perceptions and cultural attitudes (“legal cynicism”) as well as the historical-structural conditions that produced them [@Bell2017]. Research on legal cynicism has found that public perceptions of abusive police practices can reduce willingness to report crimes or cooperate with law enforcement [@Desmond2016; @Tyler2014]. Because police are the most visible agents of the state, perceptions of unfairness in policing can translate to a recognition of actual group-level structural exclusion from public institutions [@Bell2017]. New evidence related to ticketing and legal estrangement has recently emerged: in a survey of residents in three Georgia cities that rely heavily on fines and fees revenue, residents who were ticketed reported lower levels of trust in police and government compared to residents who were not ticketed [@CarpenterII2019].

Sarah Brayne found related results in her analysis of survey data---as the intensity of criminal legal contact escalates (i.e. stopped, arrested, convicted), people subjected to criminalization are increasingly likely to avoid institutions that they perceive as surveilling, such as financial or educational institutions [@Brayne2014]. She calls this phenomenon “system avoidance,” and while her analysis did not specifically assess voting behavior, her findings demonstrate how criminal legal contact can discourage Americans’ participation in public institutions as well as how the effect may vary alongside the intensity of criminal legal intervention. Similarly, @Remster2018a finds that men who return home from prison are more likely to avoid contact with their childrens’ schools.

In addition to influencing perceptions of government and participation in public institutions, recent scholarship indicates that criminalization influences political behavior. The “hidden curriculum” of the criminal legal system [@Justice2014; @Meares2017] teaches individuals about their position within the political system, and the criminal legal system consequently serves as a primary site of political socialization for convicted individuals and their families [e.g. @Lee2014].

Using existing national survey data and extended interviews, Vesla Weaver and Amy Lerman found that trust in government and willingness to vote decrease as individuals progress through increasingly intense levels of criminal legal contact (questioned by police, arrested, convicted, incarcerated) [@Weaver2010]. Weaver and Lerman also investigate the potential role of criminal legal contact in facilitating a collective political consciousness among Black Americans, or the concept of a “linked fate” where “one’s individual opportunities are linked to the entire racial group” [@Lerman2014, page 197]. Their interview data lead them to conclude that while criminal legal contact does increase perceptions of linked fate among Black respondents, “a custodial linked fate is not the building block of greater activism and a mobilized racial identity; instead, it is demobilizing”---and interview respondents who had been incarcerated reported that withdrawal from public institutions and democratic participation constituted a form of self-preservation [@Lerman2014].

Lerman and Weaver’s finding has been corroborated by administrative data: Ariel White found that short jail sentences in Harris County, Texas reduced Black defendant turnout in subsequent elections by 13 percentage points---but the effect on white turnout was not statistically significant [@White2019]. Traci Burch found that North Carolina block groups with higher concentrations of people who had experienced incarceration or criminal legal supervision saw substantial reductions in voter turnout, and confirmed her findings at the individual level [@Burch2014].

Although most work documents a uniformly demobilizing effect from contact with the criminal legal system, recent work from Hannah Walker [-@Walker2020; -@Walker2020a] complicates this narrative by arguing that, when Black Americans understand their contact with the criminal legal system in the context of a narrative of racial injustice, such contact can actually prove mobilizing. She explains: “experiences with punitive policies are subject to interpretation and, when understood through narratives of injustice, lead to wholly different participation outcomes than we usually predict” [@Walker2020, page 132]. While she finds that this sense of injustice is especially likely to increase political participation in non-voting ways (such as attending a protest or signing a petition), the logic could extend to electoral participation as well given a strong narrative or relatively weak contact with the criminal legal system.

Despite the growing body of high-quality scholarship on the democratic implications of the American criminal legal system, we still know remarkably little about the effects of the most routine sort of contact---police stops and tickets that do not result in an arrest. We leverage Walker’s novel insights about the potentiality of a narrative of injustice to re-think how these sorts of contact with the state can structure political participation.

@@

Fines and fees are situated in a distinct historical legacy of racist exploitation that may uniquely facilitate a collective “sense of injustice” among Black Americans---the sort of collective perception that @Walker2020a argues could prove to be politically mobilizing.

Cedric Robinson’s articulation of racial capitalism offers a useful theoretical framework for interpreting the trajectory of American anti-Black political economy [@Robinson2000]. Brittany Friedman theorizes fines and fees as a “progression of racial capitalism penology,” emphasizing the centrality of carceral immobility and financial capture in contemporary fines and fees practices [@Friedman2020, page 177]. Robinson argues that European industrial capitalism continued a longer political-economic trend of differentiating racial groups in order to dispossess them and exploit their labor.

Following the Civil War and the rise of global industrial capitalism, slave patrols were replaced by more formal police forces [@EqualJusticeInitiative2017; @Hinton2021]. The implementation of the Black Codes, convict leasing, and sharecropping systems marked a transition to a new stage of racial capitalism, wherein criminalization provided an easily-exploited labor supply that offset costs that the state incurred by coercing their labor [@Lichtenstein1996; @Mancini1978]. Contemporary fines and fees practices have been directly compared to sharecropping because both structures share “a long-term pattern of financial exploitation based on race… that limits the life-choices of the affected population” [@Blessett2016, page 113].

In the 20th century, residential segregation simultaneously fueled the racial wealth gap and ensured that criminalization was both spatially and racially concentrated [@Rothstein2017; @Hinton2021]. At the level of federal policy, racial threat and whites’ attribution of criminality to Blackness [@Muhammad2010] prompted a shift “from the War on Poverty to the War on Crime” wherein Black communities’ demands for material security and safety were selectively amplified to promote increased and harsher policing and incarceration [@Hinton2017; @Vargas2017a].

The 2008 financial crisis solidified the connection between fiscal austerity and policing, prompting a decline in tax revenues that devastated state and local governments---but law enforcement agencies were largely spared from cuts. At the local level, this means that police budgets today constitute about thirty percent of city budgets nationwide, and cities where residents have lower median incomes dedicate a larger proportion of spending to policing [@Friedman2020]. Similarly, municipalities where a larger share of the population is Black raise more fines and fees revenue per-capita than less-Black municipalities [@Sances2017; @UnitedStatesCommissiononCivilRights2017]. Although major urban centers collect high dollar amounts of fines and fees, some of the most egregious ticketing practices measured on a per-capita basis are concentrated in suburbs or smaller cities with municipal courts---Ferguson being a prominent example [@UnitedStatesDepartmentofJusticeCivilRightsDivision2015; @Mughan2020]. In municipalities with substantial Black populations, then, people who are most heavily affected by criminalization are expected to foot the bill for their own mistreatment. @Sances2017, however, finds that the discriminatory imposition of tickets can be mitigated by descriptive representation (i.e. Black elected officials): The positive association between cities’ reliance on fines and fees and the proportion of Black residents is significantly diminished when Black communities are represented by at least one Black legislator [@Sances2017].

Racial capitalist political economy imposes fiscal constraints regardless of city officials’ identities or intentions. In their study of fines and fees reliance in Chicago suburbs, Josh Pacewicz and John Robinson, III show how the racialization of municipal opportunity structures prevents majority-Black suburban governments from acquiring sufficient revenue from “good” sources such as sales taxes or property taxes, catalyzing a reliance on fines and fees to fund government services [@Pacewicz2020]. Pacewicz and Robinson III note that racialized municipal opportunity structures are tied to the political-historical processes of neoliberalism and fiscal austerity, an observation that aligns with recent scholarship detailing the origins of fees paid by incarcerated people [@Kirk2020].

Today, fines and fees primarily target Black Americans: traffic stops and searches disproportionately affect Black motorists nationwide [@Pierson2020], and a host of local analyses confirm the national finding [@Dunn2009; @Goncalves2020; @Mughan2020]. In criminal court proceedings, Black defendants tend to face more severe monetary sanctions [@Edwards2020; @Harris2011], and a 2020 Federal Reserve survey found that five percent of white respondents reported “unpaid legal expenses, fines, or court costs,” whereas twelve percent of Black respondents reported such debt [@BoardofGovernorsoftheFederalReserveSystem2020].

Research regarding police stops and searches in response to local fiscal crises shows how decreasing local revenues affect racist police practices. One study of all U.S. city governments found that local fiscal stress was related to increased “revenue-generating arrests” (e.g. arrests involving civil asset forfeiture) and that Black and Hispanic people were arrested at higher rates [@Makowsky2009]. A study of four million North Carolina traffic stops found that fiscal stress caused a decrease in the probability of white drivers being searched, but no decrease for Black drivers [@Shoub2020]. This suggests that higher reliance on fines and fees results in government officials and police targeting individuals who hold less political power---whether those individuals are out-of-state motorists or members of local marginalized communities [@Makowsky2009].

Black Americans can therefore situate exploitative ticketing practices in a longer historical legacy of collective disparate treatment and violence at the hands of the state. We argue that Black voters’ recognition of discriminatory ticketing practices facilitates a “linked fate” political analysis that motivates voting behavior distinct from other forms of criminalization, such as criminal conviction or incarceration. We expect both ticketing and incarceration to spark a recognition of “a sense of injustice” among Black Americans, but we argue that ticketing should ultimately mobilize Black voters (unlike incarceration, which discourages Black voters more than whites [@White2019]).

Ticketing produces substantively different experiences than being arrested or incarcerated. While tickets can lead to driver’s license suspensions or, in a smaller subset of cases, nonpayment incarceration, tickets are generally far less disruptive than criminal convictions and prison terms, and we argue that this difference should lead to different political behaviors. For example, one participant in @Lerman2014 who had served five years of probation after a criminal conviction reported, “I feel like if I contact a senator or governor, they’ll probably want to put me in jail and leave me as a troublemaker... That’s how I actually feel: ‘I better stay below the radar ’cause this guy’s tryin’ to start trouble’” [@Lerman2014, page 210]. While formerly incarcerated people or people who have experienced criminal legal supervision may avoid voting due to the same type of fear of state surveillance and control that motivates their avoidance of medical or educational institutions [@Brayne2014], we argue that being ticketed should not produce the same chilling effect---ticketing should not prompt Black people to use political withdrawal as a self-preservation strategy.

Additionally, we argue that “policing for profit”---the experience of being ticketed while a municipality is palpably ramping up its fines and fees practices to collect revenue---is also substantively distinct from the experience of being arrested or incarcerated. In doing so, we extend Epp et al.’s theory of pretextual traffic stops politically socializing Black voters to the specific context of discriminatory police ticketing [@Epp2014]. Black Americans who experience disproportionate ticketing are likely aware that they are being targeted by police for reasons unrelated to traffic safety. One Buffalo, New York resident who is a plaintiff in a lawsuit regarding racist ticketing practices told a local newspaper in 2019: “In my mind, they weren’t balancing the budgets for Black lives, but with Black lives” [@McLeod2019]. The city created a new traffic violations agency in 2015, spurring its yearly ticketing numbers to jump from around 32,000 to over 52,000. This financial aspect of ticketing should make it distinct from arrests and incarceration: much like the coercion of labor under the Black Codes, contemporary fines and fees practices are blatantly an effort to extract wealth from Black Americans. Although 20th century political rhetoric centered public safety as the mission of policing [CITE], exploitative ticketing practices cannot be reconciled with such rhetoric. In other words, to the extent that mass incarceration could be justified using narratives of “colorblind racism” (i.e., there’s a lot of crime happening in Black communities, so that’s why there’s more policing in those areas), racist ticketing practices do not enjoy the same rhetorical convenience.

Ticketing is by far the most common form of criminal legal contact; in 2018, more than 24 million people experienced police contact in the context of a traffic stop, whereas about 5.17 million people are currently disenfranchised as a result of a felony conviction [@Harrell2020; @Uggen2020]. At the same time, the most egregious local ticketing increases occurred largely over the past two decades [@Singla2020], and increases have been concentrated largely in Black municipalities [@Sances2017]. These facts have been extensively discussed in mainstream media and should be widely known among Black voters [@Eligon2019; @Ehrenfreund2014]. We expect a larger share of more affluent Black voters to be affected by ticketing than other forms of criminalization---arrests and incarceration are most common among individuals experiencing poverty [@Looney2018]. Because perceptions of “linked fate” and related critiques of policing are even more salient among members of the Black middle class [@Dawson1995], we expect Black middle class voters to be mobilized by egregious ticketing practices. In other words, we argue that ticketing should ensnare more Black voters who are more likely to politicize the issue.

>>>>

author:
- Jonathan Ben-Menachem^[PhD Student, Columbia University, Department of Sociology (jb4487@columbia.edu)]
- Kevin Morris^[PhD Student, CUNY Graduate Center, Department of Sociology (kmorris@gradcenter.cuny.edu)]
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    keep_tex: true
bibliography: "policing_turnout.bib"
link-citations: yes
fontsize: 12pt
header-includes:
    - \usepackage{rotating}
    - \usepackage{setspace}
abstract: |
  The American criminal legal system is an important site of political socialization: sociologists show how the alienating effect of criminalization pushes people away from public institutions, and political scientists have measured the effects of policing and incarceration on political participation. Despite this burgeoning literature, no research has directly investigated how police ticketing practices affect political participation. We use three approaches to answer this question: national survey data, a city-level analysis using Census of Governments data and the national voter file, and individual-level ticketing data from Hillsborough County, Florida between 2003 and 2019. The survey data show that Black people who report that they or a family member were recently stopped by police are more likely to vote than those who reported no such contact. Nationally, we estimate that each 10 percent increase in city fines and fees collections between 2012 and 2017 is associated with a roughly 0.04 percentage point turnout decrease but was not associated with Black turnout. In Tampa, Florida, we estimate that being ticketed decreases an individual's likelihood of voting in federal elections by 1.7 percentage points, but depressive effects are smaller for Black residents.

@@

Fines and fees are increasingly recognized as a form of racist revenue extraction connected to marginalized communities' alienation from government. After Michael Brown was killed by the Ferguson Police Department in 2014, a US Department of Justice investigation into the city's police and courts demonstrated that the municipality was engaged in a practice that advocates now refer to as "policing for profit." The city's reliance on fines and fees to fund government functions grew from 13 to 23 percent of the total budget between fiscal years 2012 and 2015. From 2012 to 2014, the Department of Justice found that 85 percent of vehicle stops, 90 percent of citations, and 93 percent of arrests targeted Black people. By contrast, just two-thirds of Ferguson's residents are Black [@UnitedStatesDepartmentofJusticeCivilRightsDivision2015].

It wasn't just a Ferguson problem, or even a Missouri problem. American cities' reliance on fines and fees revenue increased significantly following the 2008 recession---as local tax revenues dropped and tax increases became less politically viable, jurisdictions increased the amounts of fines and fees and imposed them more frequently in order to fund government services [@Singla2020; @Harris2017].

Work exploring how criminalization directly and indirectly influences political participation has exploded in recent years. Scholars have found that criminal legal contact (i.e. arrest, conviction, incarceration) consistently discourages voting [@Weaver2010; @Burch2011; @White2019a]. Such work has largely focused on the effects of highly disruptive contact with the criminal legal system such as incarceration and felony convictions [@Burch2014; @Lee2014]. Building on recent theoretical work from @Walker2020a and an analysis of the historical context of fees and fines collections in the United States, this paper is the first to explicitly theorize how local ticketing practices shape the political socialization of ticketed individuals. Additionally, our project represents the first use of individual-level administrative data to identify the causal effect of ticketing on voter behavior.

In addition to our theoretical contribution, the use of administrative data marks an important step forward in our understanding of how low-level contact with the criminal legal system structures political participation. Past work looking at the individual-level effects of low-level contact has relied entirely on survey or qualitative data [e.g. @Walker2014; @Weaver2010]. Existing research allows for the testing of specific psychological mechanisms and personal interpretations of criminal legal contact, but does not allow us to generalize more broadly. As Weaver and Lerman [-@Weaver2010, page 821] note, it may also introduce measurement bias. Our analysis investigates actual voting behavior following actual ticketing events, not reported voting behavior or reported exposure to police ticketing. The administrative data therefore allow us both to sidestep reporting error and to observe the behavior of the quarter-million individuals ticketed over a 6 year period---a far larger pool than even the most robust surveys.

Before turning to the administrative data, we begin with a novel analysis of 2020 national survey data regarding voters' experience with police stops. We find that Black respondents who reported personal or proximal experience with police stops in the last 12 months are more likely to vote than Black people who did not, while we find no statistically significant relationship between stops and voting among non-Black respondents.

Next, we join fees and fines data from municipalities around the country in 2012 and 2017 with a novel dataset estimating racial turnout for thousands of municipalities around the country. While we find that increased fees and fines likely cause lower turnout among non-Black Americans, we find no effect of fees and fines on Black turnout.

@@

The American criminal legal system directly shapes the American electorate and excludes Black voters. Today, more than five million Americans are legally barred from voting due to a felony conviction, and one in every sixteen Black Americans of voting age is disenfranchised [@Uggen2020]. Following the implementation of the Fifteenth Amendment in 1870, over a dozen U.S. states passed laws denying voting rights to people who were convicted of felonies. In their analysis of how racial threat catalyzes felon disenfranchisement, @Behrens2003 found that increases in the share of a state's prison population that is nonwhite drastically increases the likelihood of legislation permanently disenfranchising people convicted of felonies. As @Shofner1963 notes, felony disenfranchisement has often been tied explicitly to the increased presence of Black Americans in the electorate.

@@

Beyond de jure disenfranchisement, the spatial and racial concentration of criminalization influences affected communities' perceptions of government. Monica Bell calls this process "legal estrangement," a conceptual framework meant to capture those perceptions and cultural attitudes ("legal cynicism") as well as the historical-structural conditions that produced them [@Bell2017]. Research on legal cynicism has found that public perceptions of abusive police practices can reduce willingness to report crimes or cooperate with law enforcement [@Desmond2016; @Tyler2014]. Because police are the most visible agents of the state, perceptions of unfairness in policing can translate to a recognition of actual group-level structural exclusion from public institutions [@Bell2017]. New evidence related to ticketing and legal estrangement has recently emerged: in a survey of residents in three Georgia cities that rely heavily on fines and fees revenue, residents who were ticketed reported lower levels of trust in police and government compared to residents who were not ticketed [@CarpenterII2019].

Sarah Brayne found related results in her analysis of survey data---as the intensity of criminal legal contact escalates (i.e. stopped, arrested, convicted), people subjected to criminalization are increasingly likely to avoid institutions that they perceive as surveilling, such as financial or educational institutions [@Brayne2014]. She calls this phenomenon "system avoidance," and while her analysis did not specifically assess voting behavior, her findings demonstrate how criminal legal contact can discourage Americans' participation in public institutions as well as how the effect may vary alongside the intensity of criminal legal intervention. Similarly, @Remster2018a finds that men who return home from prison are more likely to avoid contact with their childrens' schools.

In addition to influencing perceptions of government and participation in public institutions, recent scholarship indicates that criminalization influences political behavior. The "hidden curriculum" of the criminal legal system [@Justice2014; @Meares2017] teaches individuals about their position within the political system, and the criminal legal system consequently serves as a primary site of political socialization for convicted individuals and their families [e.g. @Lee2014].

Using existing national survey data and extended interviews, Vesla Weaver and Amy Lerman found that trust in government and willingness to vote decrease as individuals progress through increasingly intense levels of criminal legal contact (questioned by police, arrested, convicted, incarcerated) [@Weaver2010]. Weaver and Lerman also investigate the potential role of criminal legal contact in facilitating a collective political consciousness among Black Americans, or the concept of a "linked fate" where "one's individual opportunities are linked to the entire racial group" [@Lerman2014, page 197]. Their interview data lead them to conclude that while criminal legal contact does increase perceptions of linked fate among Black respondents, "a custodial linked fate is not the building block of greater activism and a mobilized racial identity; instead, it is demobilizing"---and interview respondents who had been incarcerated reported that withdrawal from public institutions and democratic participation constituted a form of self-preservation [@Lerman2014].

Lerman and Weaver's finding has been corroborated by administrative data: Ariel White found that short jail sentences in Harris County, Texas reduced Black defendant turnout in subsequent elections by 13 percentage points---but the effect on white turnout was not statistically significant [@White2019]. Traci Burch found that North Carolina block groups with higher concentrations of people who had experienced incarceration or criminal legal supervision saw substantial reductions in voter turnout, and confirmed her findings at the individual level [@Burch2014].

Although most work documents a uniformly demobilizing effect from contact with the criminal legal system, recent work from Hannah Walker [-@Walker2020; -@Walker2020a] complicates this narrative by arguing that, when Black Americans understand their contact with the criminal legal system in the context of a narrative of racial injustice, such contact can actually prove mobilizing. She explains: "experiences with punitive policies are subject to interpretation and, when understood through narratives of injustice, lead to wholly different participation outcomes than we usually predict" [@Walker2020, page 132]. While she finds that this sense of injustice is especially likely to increase political participation in non-voting ways (such as attending a protest or signing a petition), the logic could extend to electoral participation as well given a strong narrative or relatively weak contact with the criminal legal system.

Despite the growing body of high-quality scholarship on the democratic implications of the American criminal legal system, we still know remarkably little about the effects of the most routine sort of contact---police stops and tickets that do not result in an arrest. We leverage Walker's novel insights about the potentiality of a narrative of injustice to re-think how these sorts of contact with the state can structure political participation.

@@

Fines and fees are situated in a distinct historical legacy of racist exploitation that may uniquely facilitate a collective "sense of injustice" among Black Americans---the sort of collective perception that @Walker2020a argues could prove to be politically mobilizing.

Cedric Robinson's articulation of racial capitalism offers a useful theoretical framework for interpreting the trajectory of American anti-Black political economy [@Robinson2000]. Brittany Friedman theorizes fines and fees as a "progression of racial capitalism penology," emphasizing the centrality of carceral immobility and financial capture in contemporary fines and fees practices [@Friedman2020, page 177]. Robinson argues that European industrial capitalism continued a longer political-economic trend of differentiating racial groups in order to dispossess them and exploit their labor.

Following the Civil War and the rise of global industrial capitalism, slave patrols were replaced by more formal police forces [@EqualJusticeInitiative2017; @Hinton2021]. The implementation of the Black Codes, convict leasing, and sharecropping systems marked a transition to a new stage of racial capitalism, wherein criminalization provided an easily-exploited labor supply that offset costs that the state incurred by coercing their labor [@Lichtenstein1996; @Mancini1978]. Contemporary fines and fees practices have been directly compared to sharecropping because both structures share "a long-term pattern of financial exploitation based on race… that limits the life-choices of the affected population" [@Blessett2016, page 113].

In the 20th century, residential segregation simultaneously fueled the racial wealth gap and ensured that criminalization was both spatially and racially concentrated [@Rothstein2017; @Hinton2021]. At the level of federal policy, racial threat and whites' attribution of criminality to Blackness [@Muhammad2010] prompted a shift "from the War on Poverty to the War on Crime" wherein Black communities' demands for material security and safety were selectively amplified to promote increased and harsher policing and incarceration [@Hinton2017; @Vargas2017a].

The 2008 financial crisis solidified the connection between fiscal austerity and policing, prompting a decline in tax revenues that devastated state and local governments---but law enforcement agencies were largely spared from cuts. At the local level, this means that police budgets today constitute about thirty percent of city budgets nationwide, and cities where residents have lower median incomes dedicate a larger proportion of spending to policing [@Friedman2020]. Similarly, municipalities where a larger share of the population is Black raise more fines and fees revenue per-capita than less-Black municipalities [@Sances2017; @UnitedStatesCommissiononCivilRights2017]. Although major urban centers collect high dollar amounts of fines and fees, some of the most egregious ticketing practices measured on a per-capita basis are concentrated in suburbs or smaller cities with municipal courts---Ferguson being a prominent example [@UnitedStatesDepartmentofJusticeCivilRightsDivision2015; @Mughan2020]. In municipalities with substantial Black populations, then, people who are most heavily affected by criminalization are expected to foot the bill for their own mistreatment. @Sances2017, however, finds that the discriminatory imposition of tickets can be mitigated by descriptive representation (i.e. Black elected officials): The positive association between cities' reliance on fines and fees and the proportion of Black residents is significantly diminished when Black communities are represented by at least one Black legislator [@Sances2017].

Racial capitalist political economy imposes fiscal constraints regardless of city officials' identities or intentions. In their study of fines and fees reliance in Chicago suburbs, Josh Pacewicz and John Robinson, III show how the racialization of municipal opportunity structures prevents majority-Black suburban governments from acquiring sufficient revenue from "good" sources such as sales taxes or property taxes, catalyzing a reliance on fines and fees to fund government services [@Pacewicz2020]. Pacewicz and Robinson III note that racialized municipal opportunity structures are tied to the political-historical processes of neoliberalism and fiscal austerity, an observation that aligns with recent scholarship detailing the origins of fees paid by incarcerated people [@Kirk2020].

Today, fines and fees primarily target Black Americans: traffic stops and searches disproportionately affect Black motorists nationwide [@Pierson2020], and a host of local analyses confirm the national finding [@Dunn2009; @Goncalves2020; @Mughan2020]. In criminal court proceedings, Black defendants tend to face more severe monetary sanctions [@Edwards2020; @Harris2011], and a 2020 Federal Reserve survey found that five percent of white respondents reported "unpaid legal expenses, fines, or court costs," whereas twelve percent of Black respondents reported such debt [@BoardofGovernorsoftheFederalReserveSystem2020].

Research regarding police stops and searches in response to local fiscal crises shows how decreasing local revenues affect racist police practices. One study of all U.S. city governments found that local fiscal stress was related to increased "revenue-generating arrests" (e.g. arrests involving civil asset forfeiture) and that Black and Hispanic people were arrested at higher rates [@Makowsky2009]. A study of four million North Carolina traffic stops found that fiscal stress caused a decrease in the probability of white drivers being searched, but no decrease for Black drivers [@Shoub2020]. This suggests that higher reliance on fines and fees results in government officials and police targeting individuals who hold less political power---whether those individuals are out-of-state motorists or members of local marginalized communities [@Makowsky2009].

Black Americans can therefore situate exploitative ticketing practices in a longer historical legacy of collective disparate treatment and violence at the hands of the state. We argue that Black voters' recognition of discriminatory ticketing practices facilitates a "linked fate" political analysis that motivates voting behavior distinct from other forms of criminalization, such as criminal conviction or incarceration. We expect both ticketing and incarceration to spark a recognition of "a sense of injustice" among Black Americans, but we argue that ticketing should ultimately mobilize Black voters (unlike incarceration, which discourages Black voters more than whites [@White2019]).

Ticketing produces substantively different experiences than being arrested or incarcerated. While tickets can lead to driver's license suspensions or, in a smaller subset of cases, nonpayment incarceration, tickets are generally far less disruptive than criminal convictions and prison terms, and we argue that this difference should lead to different political behaviors. For example, one participant in @Lerman2014 who had served five years of probation after a criminal conviction reported, "I feel like if I contact a senator or governor, they'll probably want to put me in jail and leave me as a troublemaker... That's how I actually feel: ‘I better stay below the radar 'cause this guy's tryin' to start trouble'" [@Lerman2014, page 210]. While formerly incarcerated people or people who have experienced criminal legal supervision may avoid voting due to the same type of fear of state surveillance and control that motivates their avoidance of medical or educational institutions [@Brayne2014], we argue that being ticketed should not produce the same chilling effect---ticketing should not prompt Black people to use political withdrawal as a self-preservation strategy.

Additionally, we argue that "policing for profit"---the experience of being ticketed while a municipality is palpably ramping up its fines and fees practices to collect revenue---is also substantively distinct from the experience of being arrested or incarcerated. In doing so, we extend Epp et al.'s theory of pretextual traffic stops politically socializing Black voters to the specific context of discriminatory police ticketing [@Epp2014]. Black Americans who experience disproportionate ticketing are likely aware that they are being targeted by police for reasons unrelated to traffic safety. One Buffalo, New York resident who is a plaintiff in a lawsuit regarding racist ticketing practices told a local newspaper in 2019: "In my mind, they weren't balancing the budgets for Black lives, but with Black lives" [@McLeod2019]. The city created a new traffic violations agency in 2015, spurring its yearly ticketing numbers to jump from around 32,000 to over 52,000. This financial aspect of ticketing should make it distinct from arrests and incarceration: much like the coercion of labor under the Black Codes, contemporary fines and fees practices are blatantly an effort to extract wealth from Black Americans. In other words, to the extent that mass incarceration could be justified using narratives of "colorblind racism" (i.e., high crime rates in Black communities results in more policing in those areas), racist ticketing practices do not enjoy the same rhetorical convenience. It may not be possible for revenue-motivated, racist ticketing to be reconciled with the idea that policing is primarily meant to improve public safety.

Ticketing is by far the most common form of criminal legal contact; in 2018, more than 24 million people experienced police contact in the context of a traffic stop, whereas about 5.17 million people are currently disenfranchised as a result of a felony conviction [@Harrell2020; @Uggen2020]. At the same time, the most egregious local ticketing increases occurred largely over the past two decades [@Singla2020], and increases have been concentrated largely in Black municipalities [@Sances2017]. These facts have been extensively discussed in mainstream media and should be widely known among Black voters [@Eligon2019; @Ehrenfreund2014]. We expect a larger share of more affluent Black voters to be affected by ticketing than other forms of criminalization---arrests and incarceration are most common among individuals experiencing poverty [@Looney2018]. Because perceptions of "linked fate" and related critiques of policing are even more salient among members of the Black middle class [@Dawson1995], we expect Black middle class voters to be mobilized by egregious ticketing practices. In other words, we argue that ticketing should ensnare more Black voters who are more likely to politicize the issue.

Finally, the rise of the Black Lives Matter movement in 2013 constitutes a significant development for the political socialization of Black Americans, and the movement began in municipalities that targeted Black residents with tickets. The movement garnered national attention and momentum during the Ferguson protests in 2014, where racially discriminatory ticketing practices were particularly egregious. Furthermore, local Black Lives Matter activism can change police practices. One recent study found that census places that saw Black Lives Matter protests experienced a 15 to 20 percent decrease in police killings over the following five year period [@Campbell2021]. It's possible that Black Lives Matter facilitated a more robust and prevalent linked-fate political analysis among Black Americans during the time frame of our analysis---the movement arose shortly after fines and fees became a more salient and widespread threat to Black Americans' material security, and many early Black Lives Matter protests either occurred in municipalities with exploitative ticketing practices or directly followed traffic stops that escalated into police violence [@Holpuch2015; @LaFraniere2016].

Today, fines and fees disproportionately affect Black people, and Black political mobilization appears to mitigate some of the most egregious ticketing practices [@Sances2017]. As we have shown, contemporary fees and fines practices are just one iteration of a longer anti-Black project in the United States. Moreover, organizing in recent years by groups affiliated with the BLM movement have made this narrative easily accessible to citizens around the country. Based on the strength and accessibility of this narrative and the injustice framework proposed by @Walker2020a, we  expect that Black voters' political behavior in response to police ticketing will be distinct from the political behavior of white voters, and further that ticketing should mobilize Black voter turnout in cities that have recently ramped up ticketing.
